---
title: Descriptive analyses and basic plotting in R
author: Aaron Lun, Catalina Vallejos
date: 11 December 2016
output: 
  html_document:
    fig_caption: false
---

# Introducing (R)markdown

Use triple backticks to indicate code environments:

```{r}
a <- 1
print(a)
```

Set `eval=FALSE` to show the code, but not run it:

```{r, eval=FALSE}
a <- 2
print(a)
```

Set `echo=FALSE` to run the code, but not show it:

```{r, echo=FALSE}
a <- 3
print(a)
```

Set `results="hide"` to hide the results:

```{r, results="hide"}
a <- 4
print(a)
```

For plots, set `fig.width` and `fig.height` to change the dimensions of the output plot:

```{r, fig.width=10, fig.height=6}
hist(rnorm(10000), xlab="X", main="Normal distribution")
```

By default, _knitr_ (the software that compiles the Rmarkdown files) will continue processing the file, even if an error has occurred within a code block.
To make it stop at the first error, it's advisable to set:

```{r}
knitr::opts_chunk$set(error=FALSE)
```

The same argument can also be used to globally set other arguments for all chunks, e.g., `fig.width`.
See `?knitr::opts_chunk` and http://yihui.name/knitr/options#chunk_options for more details.

# Exploring the `cars` dataset

## Initial examination of the data

Let's look at one of the datasets that come with the R installation.
From the documentation in `?cars`:

> The data give the speed of cars and the distances taken to stop.
> Note that the data were recorded in the 1920s.  

We can inspect the first few elements:

```{r}
head(cars)
```

Or the last few elements - how would we do that?

```{r}
tail(cars) 
```

We can have a look at some summary statistics:

```{r}
summary(cars)
```
    
Speed is in miles per hour (mph), while distance is in feet; this seems reasonable for cars from the 1920s.

What functions would we use to directly calculate:

- the mean?
- the median?
- the standard deviation?
- the mininimum or maximum?
- the first or third quartiles?

```{r}
mean(cars$speed) # mean 
median(cars$speed) # median 
sd(cars$speed) # standard deviation 
min(cars$speed) # minimum 
max(cars$speed) # maximum 
quantile(cars$speed, p=0.25) # first quartile 
quantile(cars$speed, p=0.75) # third quartile 
```

## Generating some summary plots

We can make histograms of the two sets of values:

```{r}
hist(cars$speed, xlab="Speed (mph)", main="Speed distribution")
hist(cars$dist, xlab="Distance (ft)", main="Distance distribution")
```

What happens if you set:

- `breaks=10`? 
- `col="red"`? 
- `xlim=c(0, 50)`?
- `border="grey"`? 

```{r}
# Smaller bars at a higher resolution: 
hist(cars$speed, xlab="Speed (mph)", main="Speed distribution", breaks=10) 
# Coloured bars: 
hist(cars$speed, xlab="Speed (mph)", main="Speed distribution", col="red") 
# Limits to the x-axis boundaries: 
hist(cars$speed, xlab="Speed (mph)", main="Speed distribution", xlim=c(0, 50)) 
# Coloured borders: 
hist(cars$speed, xlab="Speed (mph)", main="Speed distribution", col="grey") 
```

(__Note:__ how do you find out what arguments can be supplied to a function?)

We can also make a scatter plot of speeds vs distances, below.
Why is the speed on the x-axis?

```{r}
plot(cars$speed, cars$dist, xlab="Speed (mph)", ylab="Distance (ft)")
```

What happens if we set:

- `pch=16`, or `pch=4`?
- `cex=2`, or `cex.axis=2`, or `cex.label=2`?

```{r}
# Change the type of point: 
plot(cars$speed, cars$dist, xlab="Speed (mph)", ylab="Distance (ft)", pch=4) 
# Change the size of point: 
plot(cars$speed, cars$dist, xlab="Speed (mph)", ylab="Distance (ft)", cex=2) 
# Change the size of the axis text: 
plot(cars$speed, cars$dist, xlab="Speed (mph)", ylab="Distance (ft)", cex.axis=2) 
```

Let's say I want to highlight points corresponding to cars travelling above 10 mph (the speed limit at the time).
How do I select these points?
How do I highlight them on the plot (hint: `points`)?
How do I add a legend specifying the meaning of the highlights (hint: `legend`)?

```{r}
plot(cars$speed, cars$dist, xlab="Speed (mph)", ylab="Distance (ft)")
above.limit <- cars$speed > 10 
points(cars$speed[above.limit], cars$dist[above.limit], col="red", pch=16) 
legend("topleft", col=c("black", "red"), pch=c(1, 16), c("Safe", "Illegal")) 
```

# Exploring the `chickwts` dataset

## Initial examination of the data

From the documentation in `?chickwts`:

> An experiment was conducted to measure and compare the effectiveness of various feed supplements on the growth rate of chickens.

Examine the data:

```{r}
head(chickwts)
```

Looking at some summaries of the data:

```{r}
summary(chickwts)
```

Getting the number of samples for each feed type:

```{r}
table(chickwts$feed)
```

Previous `summary` computes values across all feed types - what's missing here?

## Computing summaries for each feed type

Using `dplyr`, we can calculate summary statistics for each feed group:

```{r}
library(dplyr)
by.feed.type <- chickwts %>%
  group_by(feed) %>%
  summarise(min = min(weight),
            q1 = quantile(weight, p = 0.25),
            median = quantile(weight, p = 0.5),
            mean = mean(weight),
            q3 = quantile(weight, p = 0.75),
            max = max(weight),
            sd = sd(weight),
            n = n())

by.feed.type
```

Notice the use of the special function `n()`, which counts how many observations 
there is in each group.


## Making various types of plots

### Making a boxplot

Let's make a boxplot:

```{r}
boxplot(chickwts$weight ~ chickwts$feed, ylab="Chick weight (g)")
```

Notice that we use the special notation `~` to specify that we want to plot the
_weight_ variable split by the _feed_ factor.

How can we colour each box plot separately?
How can we make them horizontal?
What do the points in the plot represent (hint: have a look at the `range` argument in `?boxplot`)?

```{r}
# Colouring the boxplots: 
feed.cols <- c("blue", "red", "green", "yellow", "purple", "pink") 
boxplot(chickwts$weight ~ chickwts$feed, ylab="Chickweight (g)", col=feed.cols) 
# Making the boxplots horizontal: 
boxplot(chickwts$weight ~ chickwts$feed, xlab="Chickweight (g)", horizontal=TRUE, las=1) 
# Extending whiskers to the data extremes: 
boxplot(chickwts$weight ~ chickwts$feed, ylab="Chickweight (g)", range=0) 
```

### Making a barplot

We can also make a barplot of the mean weights for all feed types.

```{r}
barplot(by.feed.type$mean)
```

This would be more useful if it had labels for each bar. Can you figure out
how to do this from the function's help (`?barplot`)?
Can you change the colour of each bar?
Can you make the bars horizontal?

```{r}
barplot(by.feed.type$mean, names.arg = by.feed.type$feed, ylab="Mean weight (g)") 
barplot(by.feed.type$mean, names.arg = by.feed.type$feed, ylab="Mean weight (g)", 
        col=feed.cols) # Changing the bar colour. 
barplot(by.feed.type$mean, names.arg = by.feed.type$feed, ylab="Mean weight (g)", 
        horiz=TRUE, las=1) # Using horizontal bars. 
```

__Advanced__: how do we add error bars?
Let's compute the standard error:

```{r}
sd.wts <- by.feed.type$sd
n.obs <- by.feed.type$n
se.wts <- sd.wts/sqrt(n.obs)
```

What's the difference between `sd.wts` and `se.wts`, and which one should we use?

Can you interpret what's happening in the code below?

```{r}
mean.wts <- by.feed.type$mean
x.pos <- barplot(mean.wts, ylab="Mean weight (g)", ylim=c(0, max(mean.wts)*1.1))
mean.plus.se <- mean.wts + se.wts
segments(x.pos, mean.wts, x.pos, mean.plus.se)
segments(x.pos+0.1, mean.plus.se, x.pos-0.1, mean.plus.se)
```

What does the first `segments` call do, and what happens if you take it out?
What does the second `segments` call do?
What happens if you take out the `ylim` argument?

```{r}
```

<!-- ## ANSWER
The first "segments" draws the vertical part of the error bar. 
The second "segments" draws the horizontal part of the error bar. 
The 'ylim' argument ensures that the entirety of the error bar is visible. 
--> 

# Plotting lines with the `Orange` dataset

## Initial examination of the data

This dataset examines the growth of a number of orange trees over time.

```{r}
head(Orange)
```

Having a look at some data summaries:

```{r}
summary(Orange)
```

How many trees do we have?
How many data points do we have per tree?

```{r}
table(Orange$Tree) 
```

Let's say we wanted to calculate the _mean_ and _median_ for both age and 
circumference of each tree:

```{r}
by.tree <- Orange %>%
  group_by(Tree) %>%
  summarise(mean_age = mean(age),
            mean_circumference = mean(circumference),
            median_age = median(age),
            median_circumference = median(circumference))
by.tree
```

Although this is a valid way of doing it, it can become quite tedious if we wanted 
to apply the same functions to many variables.

To help with this problem, there is a function, called `summarise_each`.
This function applies functions to more than one variable. It works like this:

```{r}
by.tree <- Orange %>%
  group_by(Tree) %>%
  summarise_each(funs(mean = mean(.), median = median(.)), 
                 age, circumference)
by.tree
```

There are three things to notice here:
* You specify what functions you want to compute inside the special `funs()` function.
* You need to use the special character `.` inside your functions, instead of the variable name.
* The function automatically adds the function and variable names together in the output.

Using `summarise_each`, how can we get summary statistics (mean, median, quartiles, 
minimum and maximum) for the age and circumference of each tree?

```{r}
by.tree <- Orange %>% 
  group_by(Tree) %>% 
  summarise_each(funs(min = min(.), 
                 q1 = quantile(., p = 0.25), 
                 median = quantile(., p = 0.5), 
                 mean = mean(.), 
                 q3 = quantile(., p = 0.75), 
                 max = max(.)), 
                 age, circumference) 
```


## Making line plots

Making plots of the data:

```{r}
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)")
```

How can we colour the points by the tree identity?
How can we change the shape of the points by the tree identity?
If we wanted axes on the log scale, how would we do it (hint: `?plot.default`)?

```{r}
tree.number <- as.integer(levels(Orange$Tree))[Orange$Tree] # To deal with factors... see later.
tree.col <- c('red', 'blue', 'grey', 'orange', 'darkgreen')
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)", 
    col=tree.col[tree.number]) # Each point coloured according to the tree. 
tree.pch <- c(1, 2, 3, 4, 5)
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)", 
    pch=tree.pch[tree.number]) # Each point has shape depending on the tree. 
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)", 
    log="xy") # Putting both axes on the log-scale.  
```

Each data point is linked to one at an earlier time, _for the same tree_ - how do we represent this (hint: `lines`)?

```{r}
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)")
is.tree.1 <- Orange$Tree==1
lines(Orange$age[is.tree.1], Orange$circumference[is.tree.1]) 
lines(Orange$age[Orange$Tree==2], Orange$circumference[Orange$Tree==2]) 
lines(Orange$age[Orange$Tree==3], Orange$circumference[Orange$Tree==3]) 
lines(Orange$age[Orange$Tree==4], Orange$circumference[Orange$Tree==4]) 
lines(Orange$age[Orange$Tree==5], Orange$circumference[Orange$Tree==5]) 
```

There is a way to automate this kind of repetitive task using a programming technique 
called a `for` loop.

Try and see if you understand what the following code does:

```{r}
plot(Orange$age, Orange$circumference, xlab="Age (days)", ylab="Circumference (mm)")

for (tree in unique(Orange$Tree)) { 
    is.tree <- Orange$Tree==tree 
    lines(Orange$age[is.tree], Orange$circumference[is.tree]) 
} 
```

__WARNING!__ Be careful of the ordering of x-coordinates when using `lines`!
How do we fix the following code (hint: `order`):

```{r}
x <- runif(50)
y <- (x-0.5)^2
plot(x, y)
lines(x, y)
plot(x, y) 
o <- order(x) 
lines(x[o], y[o]) 
```

# Further on the power of R graphics

## Overview

A lot of graphical fine-tuning can be performed using the `par` command.
Looking at `?par` reveals a **LOT** of widely applicable graphical parameters that are not listed in the documentation of individual plotting functions.

To demonstrate, let's use the `trees` data:

> This data set provides measurements of the girth, height and volume of timber in 31 felled black cherry trees. 

Look at the summary of the data:

```{r}
summary(trees)
```

## Making some pairwise plots

We'd like to make plots of each variable against the other.
How can we fit this all onto one image?

```{r, fig.width=10, fig.height=5}
par(mfrow=c(1, 3))
plot(trees$Girth, trees$Height)
plot(trees$Girth, trees$Volume)
plot(trees$Volume, trees$Height)
```

What happens if we try to add another plot?
What other layouts can you think of? 

```{r}
# Executing another "plot" will start a fresh instance of the graphics device. 
plot(trees$Girth, trees$Height) 
# Alternative layouts include: 
par(mfrow=c(2,2)) 
plot(trees$Girth, trees$Height) 
plot(trees$Girth, trees$Volume) 
plot(trees$Volume, trees$Height) 
par(mfrow=c(3,1)) 
plot(trees$Girth, trees$Height) 
plot(trees$Girth, trees$Volume) 
plot(trees$Volume, trees$Height) 
```

In an interactive session, how can we get back to a one-panel plot (hint: what happens if you run `dev.off`)?

## Messing with the margins 

What happens here?

```{r}
par(mar=c(1.1, 1.1, 1.1, 1.1))
plot(trees$Girth, trees$Height)
```

Try tinkering around with the numbers - what do they represent?
(Or you can just look at `?par`.)

```{r}
par(mar=c(5.1, 1.1, 1.1, 1.1)) # Bottom margin 
plot(trees$Girth, trees$Height) 
par(mar=c(1.1, 5.1, 1.1, 1.1)) # Left margin 
plot(trees$Girth, trees$Height) 
par(mar=c(1.1, 1.1, 5.1, 1.1)) # Top margin 
plot(trees$Girth, trees$Height) 
par(mar=c(1.1, 1.1, 1.1, 5.1)) # Right margin 
plot(trees$Girth, trees$Height) 
```

## Other ways to save R plots

An alternative to Rstudio's export is to save R plots directly to file with various graphics devices.

```{r}
pdf("my_plot.pdf")
plot(trees$Girth, trees$Height)
dev.off()
```

What do the `width` and `height` arguments do?
What happens if you specify multiple plots?

```{r}
pdf("my_plot.pdf", width=10, height=6) # This specifies the size of the plot (in). 
plot(trees$Girth, trees$Height)  
plot(trees$Girth, trees$Volume) # This makes a new page of the PDF. 
dev.off() 
```

PDFs are preferable for viewing plots in high resolution, but you can also save to PNG (e.g., if there are too many data points to store in a PDF).
Some tinkering is required to save in a decent resolution, though:

```{r}
png("my_plot.png", width=2000, height=2000, res=300, pointsize=15)
plot(trees$Girth, trees$Height)
dev.off()
```

# Session information

It's always wise to store your session information, so you know what version of R (and its packages) you ran it on.

```{r}
sessionInfo() 
```

